---
title: "Cut-through corridors identification (Simple version)"
format: gfm
editor: source
---

### Overview

Identifying cut-through routes, commonly known as rat-runs, is a crucial step in transport model development, as recommended in the Department for Transport’s guidance (TAG Unit M3.1). These routes represent diversions from main roads onto local or minor roads, often to bypass congestion or avoid traffic signals. While traditionally, the identification of such routes relies heavily on local knowledge, this may not always be readily available or consistently applied. This reproducible workflow, implemented in R, offers a data-driven approach. It leverages open data sources and network analysis techniques to systematically and efficiently identify potential cut-through corridors, even in the absence of detailed local insights. This method provides a transparent and replicable process for transport planners and modellers.

## Set-up

To begin our analysis, we first need to load the necessary R packages. These packages provide the functions required for spatial data manipulation, network analysis, and data visualization.

```{r}
#| label: load-packages
#| output: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) install.packages("remotes")
pkgs = c(
    "sf",
    "tidyverse",
    "zonebuilder",
    "tmap",
    "sfnetworks",
    "tidygraph",
    "igraph",
    "paletteer"
)
remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)

```

A quick check of the version of the packages used for this:

```{r}
sapply(pkgs,packageVersion)
```

## Select Study Area

For this demonstration, we will focus our analysis on the city of Leeds. We use the `zonebuilder` package to define our study area. This package helps in creating consistent and reproducible geographical zones for analysis.

```{r}
selected_zones <- zonebuilder::zb_zone("Leeds",n_circles = 4)

# We then create a Well-Known Text (WKT) representation of the convex hull of these zones.
# This WKT filter will be used later to select network data only within our area of interest.
zones_wkt <- selected_zones |>
  st_union() |>
  st_convex_hull() |>
  st_transform(27700) |> 
  st_as_text()
```

Next, we define assumed speed limits for different road types. In a real-world scenario, this data might come from more detailed datasets like Ordnance Survey MasterMap Highways Network or OpenStreetMap (OSM). For this example, we'll use a simplified set of speeds. We also define speeds for congested conditions, **assuming a reduction of 10 mph for Motorways, A roads, and B roads**. Similarly, congested speeds might be available from different sources and might produce better results. These speeds are converted to meters per second, which is a common unit for network analysis calculations.

```{r}
custom_wp <- tibble(
  road_function = c("Local Road", "Minor Road", "B Road", "A Road",
                    "Motorway"),
  speed_ff = round(c(20,20,30,40,70)*(0.44704),5),
  speed_cg = round(c(20,20,20,30,60)*(0.44704),5),
  )
```

## Road Network Data

The road network data used in this analysis is OS OpenRoads, a freely available dataset from Ordnance Survey [here](https://osdatahub.os.uk/downloads/open/OpenRoads). The following code downloads the dataset directly:

```{r}
if(!file.exists("00_data/oproad_gpkg_gb.zip")){
  dir.create("00_data",showWarnings = F)
  u <- "https://api.os.uk/downloads/v1/products/OpenRoads/downloads?area=GB&format=GeoPackage&redirect"
  options(timeout = 360)
  download.file(u, destfile = "00_data/oproad_gpkg_gb.zip", mode = "wb")
  unzip("00_data/oproad_gpkg_gb.zip",exdir = "00_data")
}
```

We load the road links within our defined study area (using the `zones_wkt` filter) and exclude links classified as 'access roads' as these are typically not part of through routes. We then join our custom speed information to the network data.

```{r}
selected_network <- st_read(
  "00_data/Data/oproad_gb.gpkg",
  wkt_filter = zones_wkt,
  query = "SELECT * FROM \"road_link\" WHERE road_function NOT LIKE '%access%'") |> 
  left_join(custom_wp,by = "road_function") |> 
    # With the speeds assigned, we can calculate travel times for each road segment.
    # We calculate travel times for both free-flow (speed-limit) and congested conditions.
  mutate(
    tra_time_ff = length/speed_ff, # Travel time in free-flow (seconds)
    tra_time_cg = length/speed_cg  # Travel time in congested conditions (seconds)
         )

```

Let's take a quick look at the road network we've prepared. The map below shows the different types of roads in our Leeds study area.

```{r}
#| echo: false

road_levels <- c(
  "Motorway",
  "A Road",
  "B Road",
  "Minor Road",
  "Local Road"
)

roads_1 <- selected_network|> 
  mutate(road_function = factor(road_function,
                                levels = road_levels,
                                ordered = T)) |> 
    ggplot()+
  geom_sf(aes(col = road_function,
              linewidth = road_function,
              alpha = road_function))+
  theme_void()+
  labs(col = "Road Function")+
  scale_color_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,3.5),guide = 'none')+
  scale_alpha_manual(
    values = c(1,0.8,0.75,0.7,0.3),guide = 'none'
    )+
  guides(col=guide_legend(nrow=3, byrow=TRUE))+
  theme(legend.position = "right")

roads_1
```

To perform network analysis, we need to convert our spatial lines data into a graph representation. The `sfnetworks` package is ideal for this, as it integrates `sf` spatial objects with `tidygraph` network analysis capabilities. We convert the `selected_network` into an `sfnetwork` object.

A common step in network simplification is to remove 'pseudo nodes'. These are nodes that connect only two edges and don't represent true junctions or decision points in the network. Removing them simplifies the graph and can make subsequent analyses more efficient and meaningful. The `to_spatial_smooth` function helps achieve this by merging edges connected by pseudo nodes, while also summarizing their attributes (like summing lengths and travel times). We ensure that edges are only merged if they have the same `road_function`.

```{r}
net = as_sfnetwork(selected_network)

# Simplifying the network by removing pseudo nodes
net_simplified = convert(net,
                         to_spatial_smooth,
                         summarise_attributes = list(length = "sum",
                                                     tra_time_ff = "sum",
                                                     tra_time_cg = "sum",
                                                     "first" # Keep the first value for other attributes
                                                     ),
                         require_equal = "road_function") # Only merge edges with the same road function
```

The core of our analysis involves calculating 'betweenness centrality'. This metric quantifies how often a particular road segment (edge) lies on the shortest paths between pairs of other locations in the network. A higher betweenness centrality suggests a road is more critical for connecting different parts of the network. We calculate this for two scenarios: 1. **Baseline (Free-flow):** Using travel times based on speed limits. 2. **Congested State:** Using travel times that reflect congestion on major roads.

By comparing the centrality scores between these two states, we can identify roads that become disproportionately more 'central' or 'important' when major roads are congested. These are potential candidates for cut-through routes. We use a `cutoff` of 5 minutes (300 seconds) to limit the shortest path calculations, focusing on relatively local diversions.

```{r}
# Calculate edge betweenness centrality for free-flow conditions
E(net_simplified)$centrality_ff <- edge_betweenness(
  net_simplified,
  directed = F, # Undirected graph
  weights = E(net_simplified)$tra_time_ff, # Use free-flow travel times as weights
  cutoff = 5*60 # Consider paths up to 5 minutes
  )

# Calculate edge betweenness centrality for congested conditions
E(net_simplified)$centrality_cg <- edge_betweenness(
  net_simplified,
  directed = F, # Undirected graph
  weights = E(net_simplified)$tra_time_cg, # Use congested travel times as weights
  cutoff = 5*60 # Consider paths up to 5 minutes
  )

# Now, we calculate the difference in betweenness centrality.
# A positive difference means the road segment became more central during congestion.
# We also calculate a log-transformed difference to handle potential large variations and skewness.
net_centrality <- net_simplified |> 
  activate(edges) |> 
  mutate(diff_bc = centrality_cg - centrality_ff,
         log_diff = if_else(diff_bc!=0,
                            (diff_bc/abs(diff_bc))*log(abs(diff_bc)),0 # Log transform, preserving sign
                            ))
```

Let's examine the distribution of these centrality changes, specifically for local roads. The histogram below shows the log-transformed difference in betweenness centrality. Values to the right of zero indicate local roads that gained importance (centrality) under congested conditions.

```{r}
sf_edges <- st_as_sf(net_centrality, "edges") |> 
  mutate(road_function = factor(road_function,
                                levels = road_levels,
                                ordered = T))

sf_edges |>
    filter(road_function == "Local Road") |> 
    ggplot(aes(log_diff))+
    geom_histogram(col = "white")+
    theme_minimal()+
    labs(title = "Distribution of Centrality Change on Local Roads",
         x = "Log-transformed Difference in Betweenness Centrality (Congested - Free Flow)",
         y = "Frequency")
```

```{r}
#| eval: false
#| include: false

road_levels <- c(
  "Motorway",
  "A Road",
  "B Road",
  "Minor Road",
  "Local Road"
)

roads_1 <- ggplot(sf_edges)+
  geom_sf(aes(col = road_function,
              linewidth = road_function,
              alpha = road_function))+
  theme_void()+
  labs(col = "Road Function")+
  scale_color_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,3.5),guide = 'none')+
  scale_alpha_manual(
    values = c(1,0.8,0.75,0.7,0.3),guide = 'none'
    )+
  guides(col=guide_legend(nrow=3, byrow=TRUE))+
  theme(legend.position = "right")

roads_1

ggsave(plot = roads_1,filename = "roads_1.png",dpi = 320,width = 24,height = 14,units = "cm")


roads_2 <- ggplot(sf_edges)+
  geom_sf(aes(col = road_function,
              linewidth = road_function,
              alpha = road_function))+
  theme_void()+
  labs(col = "Road Function")+
  scale_color_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,3.5),guide = 'none')+
  scale_alpha_manual(
    values = c(1,0.8,0.75,0.7,0.01),guide = 'none'
    )+
  guides(col=guide_legend(nrow=3, byrow=TRUE))+
  theme(legend.position = "right")

ggsave(plot = roads_2,filename = "roads_2.png",dpi = 320,width = 24,height = 14,units = "cm")


bar_1 <- sf_edges |> 
    st_drop_geometry() |> 
    summarise(across(length,sum),.by = road_function) |>
    mutate(length = if_else(road_function == "Local Road",0,length/1e3)) |> 
    
ggplot()+
  geom_col(aes(fill = road_function,
              x =  length,
              y = fct_rev(road_function)))+
          theme_minimal()+
    geom_vline(xintercept = 0,linewidth = 1,col = "gray10")+
    theme(legend.position = "none",panel.grid.major.y = element_blank())+
    labs(x = "Total length (km)",
         y = NULL)+
  scale_fill_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
    scale_x_continuous(limits = c(0,1600))
    


ggsave(plot = bar_1,filename = "bar_1.png",dpi = 320,width = 10,height = 6,units = "cm")

bar_2 <- sf_edges |> 
    st_drop_geometry() |> 
    summarise(across(length,sum),.by = road_function) |>
    mutate(length = length/1e3) |> 
    
ggplot()+
  geom_col(aes(fill = road_function,
              x =  length,
              y = fct_rev(road_function)))+
          theme_minimal()+
    geom_vline(xintercept = 0,linewidth = 1,col = "gray10")+
    theme(legend.position = "none",panel.grid.major.y = element_blank())+
    labs(x = "Total length (km)",
         y = NULL)+
  scale_fill_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
    scale_x_continuous(limits = c(0,1600))


ggsave(plot = bar_2,filename = "bar_2.png",dpi = 320,width = 10,height = 6,units = "cm")

```

Under congested conditions, most local roads loose *importance* as the major roads provide less connectivity. However, a portion of them become more important as they provide quicker routes for drivers, potentially indicating a cut-through corridor.

We can explore the location of such corridors. These are the links with the highest values (97th percentile).

```{r}
cut_map <- sf_edges |>
    # We are interested in the change on local roads
    mutate(log_diff = if_else(road_function %in% c("Local Road"),log_diff,NA_real_)) |>
    # Identify local roads in the top 3% of centrality increase
     mutate(log_diff_p95 = log_diff > quantile(log_diff,0.97,na.rm = T)) |> 
    ggplot()+
  geom_sf(aes(col = log_diff_p95,
              linewidth = road_function,
              alpha = log_diff_p95
              ))+
  theme_void()+
  labs(col = "Potential Cut-Through",
       title = "Potential Cut-Through Corridors on Local Roads in Leeds",
       caption = "Highlighted roads are local roads in the 97th percentile of increased betweenness centrality under congestion.")+
  scale_color_manual(values = c("gray60","dodgerblue"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,2.1),guide = 'none')+
  scale_alpha_manual(
    values = c(0.2,1),guide = 'none',na.value = 0.4
    )+
  guides(col=guide_legend(nrow=3,byrow=TRUE))+
  theme(legend.position = "right")

cut_map

ggsave(plot = cut_map,filename = "cut_map.png",dpi = 320,width = 24,height = 14,units = "cm")

```
