---
title: "Cut-through corridors identification (Simple version)"
format: gfm
editor: source
---

### Overview

Identifying cut-through routes, commonly known as rat-runs, is a crucial step in transport model development, as recommended in the Department for Transport’s guidance (TAG Unit M3.1). Traditionally, this process relies heavily on local knowledge, which may not always be accessible. This reproducible workflow implemented in R utilises open data and network analysis techniques to identify these routes efficiently, even in the absence of detailed local insights.

## Set-up

We will use the following packages for the analysis and to produce the outputs

```{r}
#| label: load-packages
#| echo: false
#| output: false

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("remotes")) install.packages("remotes")
pkgs = c(
    "sf",
    "tidyverse",
    "zonebuilder",
    "tmap",
    "sfnetworks",
    "tidygraph",
    "igraph",
    "paletteer"
)
remotes::install_cran(pkgs)
sapply(pkgs, require, character.only = TRUE)

```

We will take Leeds to exemplify the process.

```{r}
selected_zones <- zonebuilder::zb_zone("Leeds",n_circles = 4)

zones_wkt <- selected_zones |>
  st_union() |>
  st_convex_hull() |>
  st_transform(27700) |> 
  st_as_text()
```

We will assume that all roads, have speed limits based on their types. Richer datasets might provide more detailed information, for example OSM. For the congested conditions, we will assume a reduction of 10 mph for the Motorways, A roads and B roads in the network.

```{r}
custom_wp <- tibble(
  road_function = c("Local Road", "Minor Road", "B Road", "A Road",
                    "Motorway"),
  speed_ff = round(c(20,20,30,40,70)*(0.44704),5),
  speed_cg = round(c(20,20,20,30,60)*(0.44704),5),
  )
```

For the road network, we will use OS OpenRoads available [here](https://www.ordnancesurvey.co.uk/products/os-open-roads).

```{r}
selected_network <- st_read(
  "00_data/oproad_gb.gpkg",
  wkt_filter = zones_wkt,
  query = "SELECT * FROM \"road_link\" WHERE road_function NOT LIKE '%access%'") |> 
  left_join(custom_wp,by = "road_function") |> 
    # We calculate travel times for all corridors in both free-flow (speed-limit) and congested conditions (10n mph lower)
  mutate(
    tra_time_ff = length/speed_ff,
    tra_time_cg = length/speed_cg
         )

```

A quick visualisation of the data we will be using:


```{r}
#| echo: false

road_levels <- c(
  "Motorway",
  "A Road",
  "B Road",
  "Minor Road",
  "Local Road"
)

roads_1 <- selected_network|> 
  mutate(road_function = factor(road_function,
                                levels = road_levels,
                                ordered = T)) |> 
    ggplot()+
  geom_sf(aes(col = road_function,
              linewidth = road_function,
              alpha = road_function))+
  theme_void()+
  labs(col = "Road Function")+
  scale_color_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,3.5),guide = 'none')+
  scale_alpha_manual(
    values = c(1,0.8,0.75,0.7,0.3),guide = 'none'
    )+
  guides(col=guide_legend(nrow=3, byrow=TRUE))+
  theme(text = element_text(family = "Roboto Condensed"), legend.position = "right")

roads_1
```

We will produce a graph representation of the network with the `sfnetworks` package.

```{r}
net = as_sfnetwork(selected_network)

net_simplified = convert(net,
                         to_spatial_smooth,
                         summarise_attributes = list(length = "sum",
                                                     tra_time_ff = "sum",
                                                     tra_time_cg = "sum",
                                                     "first"
                                                     ),
                         require_equal = "road_function")
```

We calculate the betweenness centrality for the baseline (free flow) and the congested state. And finally, we compare them

```{r}
E(net_simplified)$centrality_ff <- edge_betweenness(
  net_simplified,
  directed = F,
  weights = E(net_simplified)$tra_time_ff,
  cutoff = 5*60)

E(net_simplified)$centrality_cg <- edge_betweenness(
  net_simplified,
  directed = F,
  weights = E(net_simplified)$tra_time_cg,
  cutoff = 5*60)

net_centrality <- net_simplified |> 
  activate(edges) |> 
  mutate(diff_bc = centrality_cg - centrality_ff,
         log_diff = if_else(diff_bc!=0,
                            (diff_bc/abs(diff_bc))*log(abs(diff_bc)),0
                            ))




```

Let's visualise the different changes in centrality for local roads

```{r}
sf_edges <- st_as_sf(net_centrality, "edges") |> 
  mutate(road_function = factor(road_function,
                                levels = road_levels,
                                ordered = T))

sf_edges |>
    filter(road_function == "Local Road") |> 
    ggplot(aes(log_diff))+
    geom_histogram(col = "white")+
    theme_minimal()
```


```{r}
#| eval: false
#| include: false

road_levels <- c(
  "Motorway",
  "A Road",
  "B Road",
  "Minor Road",
  "Local Road"
)

roads_1 <- ggplot(sf_edges)+
  geom_sf(aes(col = road_function,
              linewidth = road_function,
              alpha = road_function))+
  theme_void()+
  labs(col = "Road Function")+
  scale_color_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,3.5),guide = 'none')+
  scale_alpha_manual(
    values = c(1,0.8,0.75,0.7,0.3),guide = 'none'
    )+
  guides(col=guide_legend(nrow=3, byrow=TRUE))+
  theme(text = element_text(family = "Roboto Condensed"), legend.position = "right")

roads_1

ggsave(plot = roads_1,filename = "roads_1.png",dpi = 320,width = 24,height = 14,units = "cm")


roads_2 <- ggplot(sf_edges)+
  geom_sf(aes(col = road_function,
              linewidth = road_function,
              alpha = road_function))+
  theme_void()+
  labs(col = "Road Function")+
  scale_color_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,3.5),guide = 'none')+
  scale_alpha_manual(
    values = c(1,0.8,0.75,0.7,0.01),guide = 'none'
    )+
  guides(col=guide_legend(nrow=3, byrow=TRUE))+
  theme(text = element_text(family = "Roboto Condensed"), legend.position = "right")

ggsave(plot = roads_2,filename = "roads_2.png",dpi = 320,width = 24,height = 14,units = "cm")


bar_1 <- sf_edges |> 
    st_drop_geometry() |> 
    summarise(across(length,sum),.by = road_function) |>
    mutate(length = if_else(road_function == "Local Road",0,length/1e3)) |> 
    
ggplot()+
  geom_col(aes(fill = road_function,
              x =  length,
              y = fct_rev(road_function)))+
          theme_minimal()+
    geom_vline(xintercept = 0,linewidth = 1,col = "gray10")+
    theme(legend.position = "none",panel.grid.major.y = element_blank(),
          text = element_text(family = "Roboto Condensed"))+
    labs(x = "Total length (km)",
         y = NULL)+
  scale_fill_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
    scale_x_continuous(limits = c(0,1600))
    


ggsave(plot = bar_1,filename = "bar_1.png",dpi = 320,width = 10,height = 6,units = "cm")

bar_2 <- sf_edges |> 
    st_drop_geometry() |> 
    summarise(across(length,sum),.by = road_function) |>
    mutate(length = length/1e3) |> 
    
ggplot()+
  geom_col(aes(fill = road_function,
              x =  length,
              y = fct_rev(road_function)))+
          theme_minimal()+
    geom_vline(xintercept = 0,linewidth = 1,col = "gray10")+
    theme(legend.position = "none",panel.grid.major.y = element_blank(),
          text = element_text(family = "Roboto Condensed"))+
    labs(x = "Total length (km)",
         y = NULL)+
  scale_fill_discrete(type = paletteer_d("ggsci::default_locuszoom"))+
    scale_x_continuous(limits = c(0,1600))


ggsave(plot = bar_2,filename = "bar_2.png",dpi = 320,width = 10,height = 6,units = "cm")

```

```


Under congested conditions, most local roads loose *importance* as the major roads provide less connectivity. However, a portion of them become more important as they provide quicker routes for drivers, potentially indicating a cut-through corridor.

We can explore the location of such corridors. These are the links with the highest values (97th percentile).

```{r}
cut_map <- sf_edges |>
    mutate(log_diff = if_else(road_function %in% c("Local Road"),log_diff,NA_real_)) |>
     mutate(log_diff_p95 = log_diff > quantile(log_diff,0.97,na.rm = T)) |> 
    ggplot()+
  geom_sf(aes(col = log_diff_p95,
              linewidth = road_function,
              alpha = log_diff_p95
              ))+
  theme_void()+
  labs(col = "Potential Cut-Through")+
  scale_color_manual(values = c("gray60","dodgerblue"))+
  scale_linewidth_manual(values = 1/c(1.5,1.8,2.05,2.5,2.1),guide = 'none')+
  scale_alpha_manual(
    values = c(0.2,1),guide = 'none',na.value = 0.4
    )+
  guides(col=guide_legend(nrow=3,byrow=TRUE))+
  theme(text = element_text(family = "Roboto Condensed"), legend.position = "right")

cut_map

ggsave(plot = cut_map,filename = "cut_map.png",dpi = 320,width = 24,height = 14,units = "cm")

```
